Learning From Massive Noisy Labeled Data for Image Classification

### 2015-04-01 paper.tex
Polish the section of label noise model

\section{Label Noise Model} % (fold)
\label{sec:label_noise_model}
We target on the problem of learning a classifier from a set of images with noisy labels. To be specific, we have a noisy labeled dataset $\mathcal{D}_n=\left\{\left(\vec{x}^{(1)}, \tilde{y}^{(1)}\right), \dots, \left(\vec{x}^{(N)}, \tilde{y}^{(N)}\right)\right\}$ with $n$-th image $\vec{x}^{(n)}$ and its corresponding noisy label $\tilde{y}^{(n)} \in \{1,\dots,L\}$, where $L$ is the number of classes. Based on previous analysis of the noise types, we describe how the noisy label is generated by using a probabilistic graphical model shown in Figure~\ref{fig:noise_pgm}.

@@

On the other hand, the latent variable $\vec{z}$ representing label noise type is also an 1-of-3 binary random variable. We assign three semantic meanings to each possible state of $\vec{z}$:
\begin{enumerate}
    \item The label is noise free, \ie, $\tilde{\vec{y}}$ should be equal to $\vec{y}$
    \item The label suffers from a pure random noise, \ie, $\tilde{\vec{y}}$ can take any possible value other than $\vec{y}$
    \item The label suffers from a confusing noise, \ie, $\tilde{\vec{y}}$ can take several values that are easily confused with $\vec{y}$
\end{enumerate}

Following this assignment rule, we define the conditional probability of the noisy label by
\begin{equation} \label{eq:likelihood}
    p(\tilde{\vec{y}}|\vec{y},\vec{z}) = \begin{cases}
        \mat{I}\vec{y} & \text{if } \vec{z}_1 = 1 \\
        \frac{1}{L-1}(\mat{U} - \mat{I})\vec{y} & \text{if } \vec{z}_2 = 1 \\
        \mat{C}\vec{y} & \text{if } \vec{z}_3 = 1
    \end{cases}
\end{equation}
where $\mat{I}$ is the identity matrix, $\mat{U}$ is the unit matrix (all the elements are one), and $\mat{C}$ is a sparse stochastic matrix with $\mathrm{tr}(\mat{C})=0$ and $\mat{C}_{ij}$ denoting the confusion probability between class $i$ and $j$. Then we can derive from Figure~\ref{fig:noise_pgm} the joint distribution of $\tilde{\vec{y}}, \vec{y}$ and $\vec{z}$ conditioning on $\vec{x}$:
\begin{equation} \label{eq:joint_distribution}
  p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}) = p(\tilde{\vec{y}} | \vec{y}, \vec{z}) p(\vec{y} | \vec{x}) p(\vec{z} | \vec{x})
\end{equation}

While the object class probability distribution $p(\vec{y} | \vec{x})$ is comprehensible, the semantic meaning of $p(\vec{z} | \vec{x})$ needs extra clarification: it can be seen as a property of an image, which represents how confusing the image is. Specific to our clothes classification problem, $p(\vec{z} | \vec{x})$ can be affected by different factors, including background clutter, image resolution, the style and material of the clothes, \etc.

To illustrate the relations between noisy label and ground truth, we can derive their conditional probability from Eq~\ref{eq:joint_distribution} by
\begin{equation}
    p(\tilde{\vec{y}} | \vec{y}, \vec{x}) = \sum_{\vec{z}} p(\tilde{\vec{y}}, \vec{z} | \vec{y}, \vec{x}) = \sum_{\vec{z}} p(\tilde{\vec{y}} | \vec{y}, \vec{z}) p(\vec{z} | \vec{x})
\end{equation}
which can be interpreted as a mixture model. Given an input image $\vec{x}$, the conditional probability $p(\vec{z} | \vec{x})$ can be seen as the prior of each mixture component. This makes a key difference between our work and~\cite{sukhbaatar2014learning}, where they assume $\tilde{\vec{y}}$ is conditionally independent with $\vec{x}$ if $\vec{y}$ is given. All the images share a same noise model in~\cite{sukhbaatar2014learning}, while in our approach each data sample has its own.

\subsection{Learning the Parameters} % (fold)
\label{sub:learning_the_parameters}
We exploit two deep neural networks to model $p(\vec{y} | \vec{x})$ and $p(\vec{z} | \vec{x})$ separately. Denote the parameter set of each deep model by $\theta_1$ and $\theta_2$. Our goal is to find the optimal $\theta = \theta_1 \cup \theta_2$ that maximize the incomplete log-likelihood $\log p(\tilde{\vec{y}} | \vec{x}; \theta)$. Expectation-Maximization algorithm is used to solve this problem iteratively.

For any probability distribution $q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x})$, we can derive a lower bound of the incomplete log-likelihood by
\begin{equation}
\begin{aligned}
    \log p(\tilde{\vec{y}} | \vec{x}; \theta) &= \log \sum_{\vec{y},\vec{z}} p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta) \\
    &\geq \sum_{\vec{y},\vec{z}} q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) \log \frac{p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta)}{q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x})}
\end{aligned}
\end{equation}
\textbf{E-Step} The difference between $\log p(\tilde{\vec{y}} | \vec{x}; \theta)$ and its lower bound is the Kullback-Leibler divergence $\mathrm{KL}\left(q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) || p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta)\right)$, which is equal to zero if and only if $q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) = p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta)$. Therefore, in each iteration $t$ of the E-Step, we first compute the posterior of latent variables using current parameters $\theta^{(t)}$:
\begin{multline} \label{eq:posterior}
    p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) = \frac{p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta^{(t)})}{p(\tilde{\vec{y}} | \vec{x}; \theta^{(t)})} \\
    = \frac{p(\tilde{\vec{y}}|\vec{y},\vec{z};\theta^{(t)}) p(\vec{y}|\vec{x};\theta^{(t)}) p(\vec{z}|\vec{x};\theta^{(t)})}{\sum_{\vec{y}',\vec{z}'} p(\tilde{\vec{y}}|\vec{y}',\vec{z}';\theta^{(t)}) p(\vec{y}'|\vec{x};\theta^{(t)}) p(\vec{z}'|\vec{x};\theta^{(t)})}
\end{multline}
Then the expected complete log-likelihood can be written as
\begin{equation}
    Q(\theta; \theta^{(t)}) = \sum_{\vec{y},\vec{z}} p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \log p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta)
\end{equation}
\textbf{M-Step} Since deep neural networks are exploited to model the probability $p(\vec{y}|\vec{x};\theta_1)$ and $p(\vec{z}|\vec{x};\theta_2)$, we perform gradient ascent on $Q$:
\begin{multline} \label{eq:gradient}
    \frac{\partial Q}{\partial \theta} = \sum_{\vec{y},\vec{z}} p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta} \log p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta) \\
    = \sum_{\vec{y}} p(\vec{y} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta_1} \log p(\vec{y}|\vec{x};\theta_1) + \\
      \sum_{\vec{z}} p(\vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta_2} \log p(\vec{z}|\vec{x};\theta_2)
\end{multline}

@@

We estimate $\mat{C}$ on a relatively small dataset $\mathcal{D}_c = \{(\vec{x},\vec{y},\tilde{\vec{y}})\}_N$, where we have $N$ images with both clean and noisy label. As prior information about $\vec{z}$ is not available, we just solve the following optimization problem:
\begin{equation}
    \max_{\mat{C},\vec{z}^{(1)},\cdots,\vec{z}^{(N)}} \sum_{i=1}^N \log p(\tilde{\vec{y}}^{(i)}|\vec{y}^{(i)},\vec{z}^{(i)})
\end{equation}
Obviously, sample $i$ contributes nothing to the optimal $\mat{C}^*$ if $\vec{y}^{(i)}$ and $\tilde{\vec{y}}^{(i)}$ are equal. So that we ignore those samples and reinterpret the problem in another form by exploiting Eq~\ref{eq:likelihood}:
\begin{equation}
\begin{aligned}
\max_{\mat{C},\vec{t}} \quad & E=\sum_{i=1}^{N'} \log \alpha^{\vec{t}_i} + \log (\tilde{\vec{y}}^{(i)T} \mat{C} \vec{y}^{(i)})^{1-\vec{t}_i}\\
\text{subject to} \quad & \mat{C} \text{ is a stochastic matrix of size } L\times L \\
                    & \vec{t} \in \{0,1\}^{N'}
\end{aligned}
\end{equation}
where $\alpha=\frac{1}{L-1}$ and $N'$ is the number of remaining samples. The semantic meaning of the above formulation is that we need to assign each $(\vec{y},\tilde{\vec{y}})$ pair the optimal noise type, while finding the optimal $\mat{C}$ simultaneously.

@@

Notice that if the true label of an image is class $i$ while the noisy label is class $j$, then it can only affect the value of $\mat{C}_{ij}$. Thus each column of $\mat{C}$ can be optimized separately. Theorem 1 further shows that samples with same pair of $(\tilde{\vec{y}},\vec{y})$ share a same noise type. Then what really matters is the frequency of each of the $L\times L$ pairs $(\tilde{\vec{y}},\vec{y})$. Considering a particular column $\vec{c}$, suppose there are $M$ samples affecting this column. We can count the frequency of noisy label class 1 to $L$ as $m_1,\dots,m_L$ and might as well set $m_1\geq m_2\geq \cdots \geq m_L$. The problem is then converted to
\begin{equation}
\begin{aligned}
\max_{\vec{c},\vec{t}} \quad & E=\sum_{k=1}^{L} m_k \left(\log \alpha^{\vec{t}_k} + \log \mat{c}_k^{1-\vec{t}_k}\right)\\
\text{subject to} \quad & \vec{c} \in [0,1]^L, \sum_{k=1}^{L} \vec{c}_k = 1 \\
                    & \vec{t} \in \{0,1\}^{L}
\end{aligned}
\end{equation}

>>>>

\section{Label Noise Model} % (fold)
\label{sec:label_noise_model}
We target on the problem of learning a classifier from a set of images with noisy labels. To be specific, we have a noisy labeled dataset $\mathcal{D}_n=\left\{\left(\vec{x}^{(1)}, \tilde{y}^{(1)}\right), \dots, \left(\vec{x}^{(N)}, \tilde{y}^{(N)}\right)\right\}$ with $n$-th image $\vec{x}^{(n)}$ and its corresponding noisy label $\tilde{y}^{(n)} \in \{1,\dots,L\}$, where $L$ is the number of classes. We describe how the noisy label is generated by using a probabilistic graphical model shown in Figure~\ref{fig:noise_pgm}.

@@

On the other hand, the latent variable $\vec{z}$ representing label noise type is an 1-of-3 binary random variable. It is associated with three different semantic meanings:
\begin{enumerate}
    \item The label is noise free, \ie, $\tilde{\vec{y}}$ should be equal to $\vec{y}$
    \item The label suffers from a pure random noise, \ie, $\tilde{\vec{y}}$ can take any possible value other than $\vec{y}$
    \item The label suffers from a confusing noise, \ie, $\tilde{\vec{y}}$ can take several values that are confusing with $\vec{y}$
\end{enumerate}

Following this assignment rule, we define the conditional probability of the noisy label as
\begin{equation} \label{eq:likelihood}
    p(\tilde{\vec{y}}|\vec{y},\vec{z}) = \begin{cases}
        \mat{I}\vec{y} & \text{if } \vec{z}_1 = 1 \\
        \frac{1}{L-1}(\mat{U} - \mat{I})\vec{y} & \text{if } \vec{z}_2 = 1 \\
        \mat{C}\vec{y} & \text{if } \vec{z}_3 = 1
    \end{cases}
\end{equation}
where $\mat{I}$ is the identity matrix, $\mat{U}$ is the unit matrix (all the elements are one), and $\mat{C}$ is a sparse stochastic matrix with $\mathrm{tr}(\mat{C})=0$ and $\mat{C}_{ij}$ denoting the confusion probability between class $i$ and $j$. Then we can derive from Figure~\ref{fig:noise_pgm} the joint distribution of $\tilde{\vec{y}}, \vec{y}$ and $\vec{z}$ conditioning on $\vec{x}$:
\begin{equation} \label{eq:joint_distribution}
  p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}) = p(\tilde{\vec{y}} | \vec{y}, \vec{z}) p(\vec{y} | \vec{x}) p(\vec{z} | \vec{x})
\end{equation}

While the class label probability distribution $p(\vec{y} | \vec{x})$ is comprehensible, the semantic meaning of $p(\vec{z} | \vec{x})$ needs extra clarification: it represents how confusing the image content is. Specific to our clothing classification problem, $p(\vec{z} | \vec{x})$ can be affected by different factors, including background clutter, image resolution, the style and material of the clothes, \etc.

To illustrate the relations between noisy label and ground truth, we can derive their conditional probability from Eq~\ref{eq:joint_distribution}
\begin{equation}
    p(\tilde{\vec{y}} | \vec{y}, \vec{x}) = \sum_{\vec{z}} p(\tilde{\vec{y}}, \vec{z} | \vec{y}, \vec{x}) = \sum_{\vec{z}} p(\tilde{\vec{y}} | \vec{y}, \vec{z}) p(\vec{z} | \vec{x})
\end{equation}
which can be interpreted as a mixture model. Given an input image $\vec{x}$, the conditional probability $p(\vec{z} | \vec{x})$ can be seen as the prior of each mixture component. This makes a key difference between our work and~\cite{sukhbaatar2014learning}, where they assume $\tilde{\vec{y}}$ is conditionally independent with $\vec{x}$ if $\vec{y}$ is given. All the images share a same noise model in~\cite{sukhbaatar2014learning}, while in our approach each data sample has its own.

\subsection{Learning the Parameters} % (fold)
\label{sub:learning_the_parameters}
We exploit two CNNs to model $p(\vec{y} | \vec{x})$ and $p(\vec{z} | \vec{x})$ separately. Denote the parameter set of each deep model by $\theta_1$ and $\theta_2$. Our goal is to find the optimal $\theta = \theta_1 \cup \theta_2$ that maximize the incomplete log-likelihood $\log p(\tilde{\vec{y}} | \vec{x}; \theta)$. The Expectation-Maximization algorithm is used to solve this problem iteratively.

For any probability distribution $q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x})$, we can derive a lower bound of the incomplete log-likelihood by
\begin{equation}
\begin{aligned}
    \log p(\tilde{\vec{y}} | \vec{x}; \theta) &= \log \sum_{\vec{y},\vec{z}} p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta) \\
    &\geq \sum_{\vec{y},\vec{z}} q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) \log \frac{p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta)}{q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x})}
\end{aligned}
\end{equation}
\textbf{E-Step} The difference between $\log p(\tilde{\vec{y}} | \vec{x}; \theta)$ and its lower bound is the Kullback-Leibler divergence $\mathrm{KL}\left(q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) || p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta)\right)$, which is equal to zero if and only if $q(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}) = p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta)$. Therefore, in each iteration $t$, we first compute the posterior of latent variables using current parameters $\theta^{(t)}$:
\begin{multline} \label{eq:posterior}
    p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) = \frac{p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta^{(t)})}{p(\tilde{\vec{y}} | \vec{x}; \theta^{(t)})} \\
    = \frac{p(\tilde{\vec{y}}|\vec{y},\vec{z};\theta^{(t)}) p(\vec{y}|\vec{x};\theta^{(t)}) p(\vec{z}|\vec{x};\theta^{(t)})}{\sum_{\vec{y}',\vec{z}'} p(\tilde{\vec{y}}|\vec{y}',\vec{z}';\theta^{(t)}) p(\vec{y}'|\vec{x};\theta^{(t)}) p(\vec{z}'|\vec{x};\theta^{(t)})}
\end{multline}
Then the expected complete log-likelihood can be written as
\begin{equation}
    Q(\theta; \theta^{(t)}) = \sum_{\vec{y},\vec{z}} p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \log p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta)
\end{equation}
\textbf{M-Step} Since deep neural networks are exploited to model the probability $p(\vec{y}|\vec{x};\theta_1)$ and $p(\vec{z}|\vec{x};\theta_2)$, we perform gradient ascent on $Q$:
\begin{multline} \label{eq:gradient}
    \frac{\partial Q}{\partial \theta} = \sum_{\vec{y},\vec{z}} p(\vec{y}, \vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta} \log p(\tilde{\vec{y}}, \vec{y}, \vec{z} | \vec{x}; \theta) \\
    = \sum_{\vec{y}} p(\vec{y} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta_1} \log p(\vec{y}|\vec{x};\theta_1) + \\
      \sum_{\vec{z}} p(\vec{z} | \tilde{\vec{y}}, \vec{x}; \theta^{(t)}) \frac{\partial}{\partial \theta_2} \log p(\vec{z}|\vec{x};\theta_2)
\end{multline}

@@

We estimate $\mat{C}$ on a relatively small dataset $\mathcal{D}_c = \{(\vec{x},\vec{y},\tilde{\vec{y}})\}_N$, where we have $N$ images with both clean and noisy labels. As prior information about $\vec{z}$ is not available, we just solve the following optimization problem:
\begin{equation}
    \max_{\mat{C},\vec{z}^{(1)},\cdots,\vec{z}^{(N)}} \sum_{i=1}^N \log p(\tilde{\vec{y}}^{(i)}|\vec{y}^{(i)},\vec{z}^{(i)})
\end{equation}
Obviously, sample $i$ contributes nothing to the optimal $\mat{C}^*$ if $\vec{y}^{(i)}$ and $\tilde{\vec{y}}^{(i)}$ are equal. So that we ignore those samples and reinterpret the problem in another form by exploiting Eq~\ref{eq:likelihood}:
\begin{equation}
\begin{aligned}
\max_{\mat{C},\vec{t}} \quad & E=\sum_{i=1}^{N'} \log \alpha^{\vec{t}_i} + \log (\tilde{\vec{y}}^{(i)T} \mat{C} \vec{y}^{(i)})^{1-\vec{t}_i}\\
\text{subject to} \quad & \mat{C} \text{ is a stochastic matrix of size } L\times L \\
                    & \vec{t} \in \{0,1\}^{N'}
\end{aligned}
\end{equation}
where $\alpha=\frac{1}{L-1}$ and $N'$ is the number of remaining samples. The semantic meaning of the above formulation is that we need to assign each $(\vec{y},\tilde{\vec{y}})$ pair the optimal noise type, while finding the optimal $\mat{C}$ simultaneously.

@@

Notice that if the true label is class $i$ while the noisy label is class $j$, then it can only affect the value of $\mat{C}_{ij}$. Thus each column of $\mat{C}$ can be optimized separately. Theorem 1 further shows that samples with same pair of $(\tilde{\vec{y}},\vec{y})$ share a same noise type. Then what really matters is the frequency of each of the $L\times L$ pairs $(\tilde{\vec{y}},\vec{y})$. Considering a particular column $\vec{c}$, suppose there are $M$ samples affecting this column. We can count the frequency of noisy label class 1 to $L$ as $m_1,\dots,m_L$ and might as well set $m_1\geq m_2\geq \cdots \geq m_L$. The problem is then converted to
\begin{equation}
\begin{aligned}
\max_{\vec{c},\vec{t}} \quad & E=\sum_{k=1}^{L} m_k \left(\log \alpha^{\vec{t}_k} + \log \mat{c}_k^{1-\vec{t}_k}\right)\\
\text{subject to} \quad & \vec{c} \in [0,1]^L, \sum_{k=1}^{L} \vec{c}_k = 1 \\
                    & \vec{t} \in \{0,1\}^{L}
\end{aligned}
\end{equation}
